---
title: "Nanopore Direct RNA Seq - Practical 2"
author: "Anton Enright & Jack Monahan"
date: \'`r format(Sys.time(), '%d %B, %Y')`\'
always_allow_html: yes
output:
  github_document:
    toc: true
---


```{r setup, results='hide',message=F, warning=F, echo=F}
require(knitr)
opts_knit$set(root.dir = '~/Desktop/course_data/wtac_nanopore_drna_2019')
opts_chunk$set(dev="png", 
               dev.args=list(type="cairo"),
               comment='           ##',
               dpi=300)
```

# Analysis of Nanopore Direct RNA counts


First we should change directory to where the data is

```{r}
setwd("~/Desktop/course_data/wtac_nanopore_drna_2019")
```

# Direct RNA Seq Discussion

We have 6 mouse datasets, with 3 wildtypes and 3 scrambles

We want to look for any differential gene expression (DGE) and differential transcript usage (DTU) between them.


## Preparation
We first load the R/BioConductor libraries that we need.
```{r, results='hide',message=F, warning=F}
suppressMessages(library(ggplot2)) # for plotting
suppressMessages(library(gplots)) # for plotting
suppressMessages(library(dplyr))
suppressMessages(library(RColorBrewer))

suppressMessages(library(GenomicFeatures)) #to create txdb from gtf

suppressMessages(library(tximport)) #import salmon counts
suppressMessages(library(readr)) #speeds up tximport

suppressMessages(library(DRIMSeq)) #for DTU
suppressMessages(library(DEXSeq)) #for DTU
suppressMessages(library(stageR)) #for DTU

suppressMessages(library(DESeq2)) #for DGE
```

#Modified from https://github.com/nanoporetech/pipeline-transcriptome-de & https://f1000research.com/articles/7-952/v3

```{r inputs}

colData = read.delim("datasets/samples.txt", stringsAsFactors = F)
colData$condition = as.factor(colData$condition)
rownames(colData) = colData$sample_id
samples = colData$sample_id

conds = colData$condition

#Salmon transcript counts 
salmon_files = file.path("datasets/salmon", samples, "quant.sf")
print(salmon_files)

#Transcript annotation
annotation_file = "references/mm10.96.gtf"

hmcol = colorRampPalette(brewer.pal(9, "GnBu"))(100)
cond_colours = brewer.pal(length(levels(conds)),"Set2")[conds]
names(cond_colours)=conds

```

#Make txdb from gencode transcript annotation gtf
```{r create_txdb}
#Already run
txdb = makeTxDbFromGFF(annotation_file, organism = "Mus Musculus",dataSource="Ensembl")
saveDb(txdb, "references/mm10.96.annotation.sqlite")

```

#Load txdb and generate Ensembl transcript id to gene id table
```{r txdb}
txdb = loadDb("references/mm10.96.annotation.sqlite")
cols <- c("TXNAME", "GENEID")
tx2gene = select(txdb, keys(txdb,"TXNAME"), columns=cols, keytype="TXNAME")

#Strip version numbers from transcript ids
tx2gene$TXNAME = sapply(strsplit(tx2gene$TXNAME, ".", fixed = T), "[[", 1)
#Strip version numbers from gene ids
tx2gene$GENEID = sapply(strsplit(tx2gene$GENEID, ".", fixed = T), "[[", 1)

enolase_ids  = c("ENSMUSG00000004267","ENSMUSG00000048029","ENSMUSG00000060600","ENSMUSG00000063524")
#Remove Enolase
tx2gene = tx2gene[!(tx2gene$GENEID %in% enolases),]

```

#Import Salmon transcript counts
```{r import_salmon}
#Import salmon transcript counts
#Don't use TPMs, use NumReads i.e. txi.salmon$counts
txi.salmon = tximport(salmon_files, type = "salmon", txOut = TRUE)
colnames(txi.salmon$abundance) = samples
colnames(txi.salmon$counts) = samples

```


#Analysis of Differential Transcript Usage
```{r Filter}

#Filter out unexpressed transcripts
keep_feature = rowSums(txi.salmon$counts) > 0
transcript_counts = txi.salmon$counts[keep_feature,]
transcript_counts = as.data.frame(transcript_counts)

#Filter transcript counts with DRIMSeq
#Make data.frame with the gene ID, the feature (transcript) ID, and then columns for each of the samples
transcript_counts$TXNAME = rownames(transcript_counts)
transcript_counts = merge(tx2gene, transcript_counts, by = "TXNAME")
colnames(transcript_counts)[c(1,2)] = c("feature_id","gene_id")

d = dmDSdata(counts=transcript_counts, samples=colData)
counts_unfiltered <- counts(d)

# DRIMSeq filter parameters
min_samps_gene_expr = 2 # or 3 normally
min_samps_feature_expr = 1
min_gene_expr = 10 
min_feature_expr =  3

d = dmFilter(d, min_samps_gene_expr = min_samps_gene_expr, min_samps_feature_expr = min_samps_feature_expr, min_gene_expr = min_gene_expr, min_feature_expr = min_feature_expr)
counts_filtered = counts(d)

```


```{r DEXSeq}
design <- model.matrix(~condition, data=DRIMSeq::samples(d))

#DEXSeq (for differential exon usage)
sample.data = DRIMSeq::samples(d)
trans_counts.data = round(as.matrix(counts_filtered[,-c(1:2)]))
dxd = DEXSeqDataSet(countData=trans_counts.data, sampleData=sample.data, design= ~sample + exon + condition:exon, featureID=counts(d)$feature_id, groupID=counts(d)$gene_id)
dxd = estimateSizeFactors(dxd)
dxd = estimateDispersions(dxd, fitType='local')
dxd = testForDEU(dxd, reducedModel=~sample + exon)
dxd = estimateExonFoldChanges( dxd, fitExpToVar="condition")
dxr = DEXSeqResults(dxd, independentFiltering=FALSE)

#pdf("analyses/results_dtu.pdf")
plotMA(dxr, cex=0.8, alpha=0.05) 
plotDispEsts(dxd)
#dev.off()

qval = perGeneQValue(dxr) 
dxr.g = data.frame(gene=names(qval), qval)
dxr.g = dxr.g[order(dxr.g$qval),]

dxr = as.data.frame(dxr[,c("featureID", "groupID", "pvalue")])
dxr = dxr[order(dxr$pvalue),]

write.table(dxr.g, file="analyses/results_dtu_gene.txt", sep="\t", quote = F)
write.table(dxr, file="analyses/results_dtu_transcript.txt", sep="\t", quote = F)

# stageR analysis on differential transcript usage results
pConfirmation <- matrix(dxr$pvalue, ncol=1)
dimnames(pConfirmation) <- list(dxr$featureID, "transcript")
pScreen <- qval
tx2gene <- as.data.frame(dxr[,c("featureID", "groupID")])

stageRObj <- stageRTx(pScreen=pScreen, pConfirmation=pConfirmation, pScreenAdjusted=TRUE, tx2gene=tx2gene)
# note: the choice of 0.05 here means you can *only* threshold at 5% overall false discovery rate (OFDR) later
stageRObj <- stageWiseAdjustment(stageRObj, method="dtu", alpha=0.05)
suppressWarnings({dex.padj <- getAdjustedPValues(stageRObj, order=FALSE, onlySignificantGenes=FALSE)})

write.table(dex.padj, file="analyses/results_dtu_stageR.txt", sep="\t", quote = F)

```

```{r plot_DTU}

# stageR results:
stageR = dex.padj
names(stageR) = c("gene_id", "transcript_id", "p_transcript", "p_gene")

# Read filtered counts:
counts = counts_filtered
names(counts)[2]<-"transcript_id"

# Join counts and stageR results:
df = merge(counts, stageR, by = c("gene_id", "transcript_id"))
df <- df[order(df$p_gene),]

#Drop columns
scols = setdiff(names(df),c("gene_id", "transcript_id", "p_gene", "p_transcript"))

# Normalise counts by the library size (i.e. the total counts per sample)
for(sc in scols){
    df[sc] <- df[sc] / sum(df[sc])
}

# Melt data frame:
tdf <- df %>% gather(key='sample', value='norm_count',-gene_id, -transcript_id, -p_gene, -p_transcript)

# Add sample group column:
sampleToGroup<-function(x){
    return(colData[x,]$condition)
}

tdf$group <- sampleToGroup(tdf$sample)

# Filter for significant genes:
sig_level <- 0.05
genes <- as.character(tdf[which(tdf$p_gene < sig_level),]$gene_id)
genes <- unique(genes)

pdf("analyses/dtu_plots.pdf")

for(gene in genes){
    gdf = tdf[which(tdf$gene_id==gene),]
    p_gene = unique(gdf$p_gene)
    p = ggplot(gdf, aes(x=transcript_id, y=norm_count)) + geom_bar(stat="identity", aes(fill=sample), position="dodge")
    p = p + facet_wrap(~ group) + coord_flip()
    p = p + ggtitle(paste(gene," : p_value=", p_gene, sep=""))
    print(p)
}

dev.off()

```

# Analysis of Differential Gene Expression
```{r DGE}

#Get gene counts from transcript counts i.e. txi.salmon$counts
txi.sum = summarizeToGene(txi.salmon, tx2gene = tx2gene, countsFromAbundance = "no")

#DESeq2 dds object from txi gene object
dds = DESeqDataSetFromTximport(txi.sum, colData = colData, design = ~condition)

#Filter out unexpressed genes
keep_feature = rowSums(counts(dds)) > 0
dds = dds[keep_feature,]

#Normal DESeq2 DGE analysis from here...

```
